Detailed structure of phospholipid bilayers with negatively charged lipids and salt concentrations from molecular dynamics simulations including the effects of electronic polarization

### 2019-05-17 Manuscript/manuscript.tex
Edits in results section PC-PS and counterions and additional nacl, kcl

In simulations, binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,NMRlipidsIV}.
For instance, the affinity of sodium counterions to a POPS bilayer 
is stronger in Lipid17/Dang simulations 
than in simulations with ECC-POPS (Fig.~\ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC} for the labeling) \cite{catte16,melcr18,NMRlipidsIV},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Although the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,NMRlipidsIV}.
Following our previous work \cite{NMRlipidsIV}, we evaluate the counterion binding affinity
to a lipid bilayer containing PS in simulations using two approaches.

@@

First, we monitor the response of POPC headgroup order parameters to the increasing amount of POPS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments, because the headgroup P--N dipole
tilts more parallel to the membrane interface \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-POPS simulation with \ce{Na+} counterions, but not in Lipid17/Dang ion simulations with \ce{Na+} counterions,
where the order parameters are unchanged or decrease and the P--N vector tilts more perpendicular to the membrane interface.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:POPS-counterions-dens}), which
compensates the negative charge of PS lipids at the interface.

Second, we monitor the response of PC headgroup order parameters to the additional
amount of sodium and potassium in POPC:POPS (5:1) mixtures.
In experiments, the addition of K$^+$ ions led to a very modest decrease of the order parameters,
while the decrease was more pronounced with Li$^+$ ions and strongest with divalent Ca$^{2+}$ and Mg$^{2+}$ ions \cite{roux90}.
According to the electrometer concept, the decrease of PC lipid head group order parameters
correlates with the amount of bound cations \cite{akutsu81,altenbach84,seelig87,roux90,catte16,NMRlipidsIV},
thus the binding affinity of the ions to POPC:POPS (5:1) mixtures increase in order K$^{+}$ $<$ Li$^{+}$  $<$ Mg$^{2+}$  $<$ Ca$^{2+}$.
The response of PC headgroup order parameters to the increasing amount of KCl is close to experiments
and the change in P--N vector is negligible in both force fields (Fig. \ref{fig:delta_ordPar_KCl_PCPS}), athought the change of the $\alpha$-carbon
order parameter may be slightly overestimated by the Lipid17/Dang simulations. This could be
explained by the slightly deeper penetration of potassium into the bilayer (Fig. \ref{fig:POPS-counterions-dens}).
Because experimental data with the additional amount of sodium in POPC:POPS (5:1) mixture was not available,
we compared the simulation results to the experimental data with the additional amounts of
lithium and potassium ions, which should give the lower and upper bounds to the sodium binding
affinity (Fig. \ref{fig:delta_ordPar_NaCl_PCPS}). The PC headgroup response to the additional
amount of sodium is clearly overestimated by the Lipid17/Dang simulation and
the response in ECC-POPS simulations is probably too close to the experimental results for lithium.
In conclusion, the results suggest that the sodium binding affinity (Fig. \ref{fig:POPS-counterions-dens}) and
the tilting of P--N vector more perpendicular to the membrane surface (Fig. \ref{fig:delta_ordPar_NaCl_PCPS})
are clearly overestimated in Lipid17/Dang simulations and slightly overestimated in ECC-POPS simulations,
while the binding affinity of potassium is better described by both force fields.

%Generally improved behaviour 
%of the POPC and POPS head group order parameters 
%with \ce{NaCl} or \ce{KCl} concentrations 
%was achieved through the combination of models 
%ECC-lipids and ECC-ions \citep{martinek17, kohagen16, Pluharova2014}. 
%Simulations with these models 
%reveal a good agreement with the NMR experiments 
%for both neutral and negatively charged membranes.
%The results are plotted for \ce{NaCl} in Fig. \ref{fig:delta_ordPar_NaCl_PCPS}, 
%and for \ce{KCl} in Fig. \ref{fig:delta_ordPar_KCl_PCPS}.

%The order parameters of the PS head groups including their changes with the content of PC lipids
%are challenging to capture correctly in classical MD simulation models. \cite{NMRlipidsIV}
%From such a perspective,
%the changes of the PS head group order parameters from ECC-lipids 
%with \ce{Na+}, and also \ce{K+}, counterions 
%agree comparably well to the experimental measurements
%(Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}). 
%Interestingly enough, 
%although the Lipid17 model performed well with \ce{K+} counterions 
%for the PC head group order parameter changes,
%it does not agree in the equivalent changes in PS head groups
%pointing at non-negligible effects of even relatively weakly binding cations 
%on the structure of negatively charged lipids.

%Despite the great improvements in the overall structure of ECC-POPS 
%and its response to PC lipids and counterions,
%there is still a large room for improving it
%to agree with the experimental measurements.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{NaCl} concentration from ECC-lipids and Lipid17/Dang simulations at 298 K.
    The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    Because experimental data with \ce{NaCl} is not available for POPC, the data for KCl and \ce{LiCl} (dashed line, left)
    are shown as lower and upper bounds for the response to \ce{NaCl}, respectively.
    The magnitudes of experimental order parameters are from Ref. \citenum{roux90} and signs from Ref. \citenum{ferreira16}.
    %which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    %The orientation of the \ce{COO^-} group is defined as 
    %the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS}
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors 
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{KCl} concentration from ECC-lipids and Lipid17/Dang simulations 
    compared with experimental values from Ref. \citenum{roux90} (signs from Ref. \citenum{ferreira16}) at 298 K.
    The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}

% This§: 
% Other models do not yield good reponse, but
% ECC-lipid describe the changes well
% => Elect. polar. necessarry.
% 
The changes of POPS headgroup order parameters are almost negligible in experiments with increasing POPC fraction in bilayer
or monovalent ion concentration in solvent
(Figs. \ref{fig:delta_ordPar_NaCl_PCPS}, \ref{fig:delta_ordPar_KCl_PCPS} and \ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
%while calcium leads to relatively large changes with low concentrations followed by a rapid saturation below 100mM (Fig. \ref{fig:delta_ordPar_CaCl_PCPS}) \cite{scherer87,roux90}.
%The interaction with \ce{K^+}, which binds very weakly to both neutral and negatively charged membranes, 
%renders a qualitatively different response of the order parameter $S^\beta$ in POPS 
%in the mixed negatively charged bilayers compared to the neutral bilayers. 
%While the order parameter $S^\beta$ increases for both \ce{Na^+} and \ce{Ca^{2+}},
%it decreases in the presence of \ce{K^+}
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}, and
%\ce{CaCl2}: Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
In Lipid17/Dang simulations, the changes of POPS headgroup order parameters with increasing amount of POPC
are overestimated, which is also related to the changes in the orientations of the P--N and C$_{\beta}$--C$_{\gamma}$
vectors (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix}). The overestimation of the order parameter changes
upon increasing amount of POPC in a bilayer 
was also observed in previous simulations with other force fields \cite{NMRlipidsIV}, but not in
ECC-POPS simulation, where also the changes in the PS headgroup orientation are more modest (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
The POPS headgroup orientation and order parameters are more sensitive also to the additional amount
of sodium and potassium in Lipid17/Dang simulations, while dependence is weaker and closer to experiments
in ECC-POPS simulations (Figs. \ref{fig:delta_ordPar_NaCl_PCPS} and \ref{fig:delta_ordPar_KCl_PCPS}).
%In contrast, ECC-lipids with ECC-ions capture the different response 
%of the order parameters $S^{\beta}$, $S^{\alpha _1}$ and $S^{\alpha _2}$ in POPS 
%to various concentrations of \ce{NaCl} and \ce{KCl} in a good agreement with the experiments
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}). 
%Such a dramatic improvement in the structural parameters 
%demonstrates that including electronic polarization in phospholipids is crucial 
%even for interactions with soft cations like \ce{K^+}.

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can
be estimated by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01 \mathrm{nm}^{-2}$ for 1 M sodium in pure ECC-POPC
simulation was reported in the previous work \cite{melcr18}.
The presence of $\sim$17\$ POPS in POPC bilayer increases the
%relative surface excess of sodium with respect to water from negative
value to $\Gamma ^{w} _{\rm Na}=0.092 \pm 0.005 \mathrm{nm}^{-2}$ 
for 0.621 M sodium, while value for potassium remains negative,
$\Gamma ^{w} _{\rm K}=-0.123 \pm 0.005 \mathrm{nm}^{-2}$ in POPC:POPS (5:1) mixture (Fig. \ref{fig:nacl-dens_PCPS}).
As discussed above, the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipids simulations. The response of headgroup order parameters to the addition of NaCl are
slightly overestimated also in ECC-POPC simulations (Fig 3 in Ref. \citenum{melcr18}),
suggesting that similar small overestimation of sodium binding may also occur to a pure POPC bilayer.
Therefore, the overestimation may arise from the sodium interactions with PC headgroup and
increase in binding affinity due to PS lipids would not be affected.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.

>>>>

In simulations, binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,NMRlipidsIV}.
For instance in the present work the affinity of sodium counterions to a POPS bilayer 
is stronger in Lipid17/Dang simulations 
than in simulations with ECC-POPS (Fig.~\ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC} for the labeling) \cite{catte16,melcr18,NMRlipidsIV},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Although the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,NMRlipidsIV}.
Following our previous work \cite{NMRlipidsIV}, we evaluate the counterion binding affinity
to a lipid bilayer containing PS in simulations using two approaches.

@@

First, we monitor the response of POPC headgroup order parameters to the increasing amount of POPS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments, because the headgroup P--N dipole
tilts more parallel to the membrane interface \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-POPS simulation with \ce{Na+} counterions, but not in Lipid17/Dang ion simulations with \ce{Na+} counterions,
where the order parameters are unchanged or decrease and the P--N vector tilts more perpendicular to the membrane interface.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:POPS-counterions-dens}), which
compensates the negative charge of PS lipids at the interface.
In contrast, ECC-lipids represent this trend correctly
with only slightly underestimated slopes of the order parameters 
when \ce{Na+} counterions are used (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).

Second, we monitor the response of PC headgroup order parameters to the additional
amount of sodium and potassium in POPC:POPS (5:1) mixtures.
In experiments, the addition of K$^+$ ions led to a very modest decrease of the order parameters,
while the decrease was more pronounced with Li$^+$ ions and was the strongest with divalent Ca$^{2+}$ and Mg$^{2+}$ ions \cite{roux90}.
According to the electrometer concept, the decrease of PC lipid head group order parameters
correlates with the amount of bound cations \cite{akutsu81,altenbach84,seelig87,roux90,catte16,NMRlipidsIV},
thus the binding affinity of the ions to POPC:POPS (5:1) mixtures increase in order K$^{+}$ $<$ Li$^{+}$  $<$ Mg$^{2+}$  $<$ Ca$^{2+}$.
The response of PC headgroup order parameters to the increasing amount of KCl is close to experiments
and the change in P--N vector is negligible in both force fields (Fig. \ref{fig:delta_ordPar_KCl_PCPS}), athough the change of the $\alpha$-carbon
order parameter may be slightly overestimated by the Lipid17/Dang simulations. 
This could be explained by the slightly deeper penetration of potassium into the bilayer (Fig. \ref{fig:POPS-counterions-dens}).

Because experimental data with the additional amount of sodium in POPC:POPS (5:1) mixture was not available,
we compared the simulation results to the experimental data with the additional amounts of
lithium and potassium ions, which should give the lower and upper bounds to the sodium binding
affinity (Fig. \ref{fig:delta_ordPar_NaCl_PCPS}). The PC headgroup response to the additional
amount of sodium is clearly overestimated by the Lipid17/Dang simulation and
the response in ECC-lipids simulations is probably too close to the experimental results for lithium.
% The following conclusion is already in the previous sentence, so I commented it out
%In conclusion, the results suggest that the sodium binding affinity (Fig. \ref{fig:POPS-counterions-dens}) and
%the tilting of P--N vector more perpendicular to the membrane surface (Fig. \ref{fig:delta_ordPar_NaCl_PCPS})
%are clearly overestimated in Lipid17/Dang simulations and slightly overestimated in ECC-POPS simulations,
%while the binding affinity of potassium is better described by both force fields.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{NaCl} concentration from ECC-lipids and Lipid17/Dang simulations at 298 K.
    % No translation in y-axis done in this plot:
    %The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    %with the same value for both order parameters such that the lower order
    %parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    Because experimental data with \ce{NaCl} is not available for POPC, the data for KCl and \ce{LiCl} (dashed line, left)
    are shown as lower and upper bounds for the response to \ce{NaCl}, respectively.
    The magnitudes of experimental order parameters are from Ref. \citenum{roux90} and signs from Ref. \citenum{ferreira16}.
  }
  \end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS}
    Changes of the head group order parameters, and the angles of P--N and C$_\beta$--C$_\gamma$ (stars) vectors 
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{KCl} concentration from ECC-lipids and Lipid17/Dang simulations 
    compared with experimental values from Ref. \citenum{roux90} (signs from Ref. \citenum{ferreira16}) at 298 K.
    % No translation in y-axis done in this plot:
    %The y-axis for the $\alpha$-carbon results of POPS (middle right) is transferred
    %with the same value for both order parameters such that the lower order
    %parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}

The changes of POPS headgroup order parameters are almost negligible in experiments with increasing POPC fraction in bilayer
or monovalent ion concentration in solvent
(Figs. \ref{fig:delta_ordPar_NaCl_PCPS}, \ref{fig:delta_ordPar_KCl_PCPS} and \ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
%while calcium leads to relatively large changes with low concentrations followed by a rapid saturation below 100mM (Fig. \ref{fig:delta_ordPar_CaCl_PCPS}) \cite{scherer87,roux90}.
%The interaction with \ce{K^+}, which binds very weakly to both neutral and negatively charged membranes, 
%renders a qualitatively different response of the order parameter $S^\beta$ in POPS 
%in the mixed negatively charged bilayers compared to the neutral bilayers. 
%While the order parameter $S^\beta$ increases for both \ce{Na^+} and \ce{Ca^{2+}},
%it decreases in the presence of \ce{K^+}
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}, and
%\ce{CaCl2}: Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
In Lipid17/Dang simulations, the changes of POPS headgroup order parameters with increasing amount of POPC
are overestimated, which is also related to the changes in the orientations of the P--N and C$_{\beta}$--C$_{\gamma}$
vectors (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix}). The overestimation of the order parameter changes
upon increasing amount of POPC in a bilayer 
was also observed in previous simulations with other force fields \cite{NMRlipidsIV}, but not in
ECC-POPS simulation, where also the changes in the PS headgroup orientation are more modest (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
The POPS headgroup orientation and order parameters are more sensitive also to the additional amount
of sodium and potassium in Lipid17/Dang simulations, while dependence is weaker and closer to experiments
in ECC-POPS simulations (Figs. \ref{fig:delta_ordPar_NaCl_PCPS} and \ref{fig:delta_ordPar_KCl_PCPS}).
These structural changes can relate to variations in the interactions between PC and PS head groups 
and also to the changing amount of counterions and PS in the composition. 
%In contrast, ECC-lipids with ECC-ions capture the different response 
%of the order parameters $S^{\beta}$, $S^{\alpha _1}$ and $S^{\alpha _2}$ in POPS 
%to various concentrations of \ce{NaCl} and \ce{KCl} in a good agreement with the experiments
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}). 
%Such a dramatic improvement in the structural parameters 
%demonstrates that including electronic polarization in phospholipids is crucial 
%even for interactions with soft cations like \ce{K^+}.

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can
be estimated by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01 \mathrm{nm}^{-2}$ for 1~M sodium in pure ECC-POPC
simulation was reported in the previous work \cite{melcr18}.
The presence of $\sim$17\$ POPS in POPC bilayer increases the
%relative surface excess of sodium with respect to water from negative
value to $\Gamma ^{w} _{\rm Na}=0.092 \pm 0.005 \mathrm{nm}^{-2}$ 
for 0.621 M sodium, while value for potassium remains negative,
$\Gamma ^{w} _{\rm K}=-0.123 \pm 0.005 \mathrm{nm}^{-2}$ in POPC:POPS (5:1) mixture (Fig. \ref{fig:nacl-dens_PCPS}).
As discussed above, the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipids simulations. The response of headgroup order parameters to the addition of NaCl are
slightly overestimated also in ECC-POPC simulations (Fig 3 in Ref. \citenum{melcr18}),
suggesting that similar small overestimation of sodium binding may also occur to a pure POPC bilayer.
Therefore, the overestimation may arise from the sodium interactions with PC headgroup and
increase in binding affinity due to PS lipids would not be affected.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.
